---
title: "AFL Analysis"
output: 
  html_document:
    code_folding: hide
editor_options: 
  chunk_output_type: inline
runtime: shiny
---

# Introduction

To keep up with the latest fixtures, results, rounds and player statistics I have made this R Notebook to do just that. Here, I have downloaded, scraped, ingested, cleaned, and analysed various pieces of AFL data. The following provides a summary of the data I have gathered, along with the following links:

-   2021 Fixtures (<https://fixturedownload.com/sport/aussie-rules>)

-   2021 Ladder (<https://afltables.com/afl/seas/2021.html> - scroll to bottom)

-   2021 Player Statistics (<https://afltables.com/afl/stats/2021a.html>)

-   Historical Results (<https://www.aussportsbetting.com/data/historical-afl-results-and-odds-data/>)

Please note that it may take a day or two after the completion of the round for the player statistics and ladder to update.

# Setup

```{r Load Packages, message=FALSE}
# Load Packages
library(rvest)
library(reshape) # must use before tidyverse due to conflict with `rename()` in dplyr
library(tidyverse)
library(janitor)
library(glue)
library(readxl)
library(writexl)
library(lubridate)
library(plotly)
library(shiny)
library(DT)
```

```{r Setup}
# Defining paths and setting working directory
home_path <- "/Users/jakekeenan/Data_Science_Projects/Not_Github/RStudio/AFL_Analysis"
fixtures_name <- "team_fixtures.xlsx"
fixtures_dir <- "Fixtures_xlsx"
setwd(home_path)
```

<!-- ### Helper Functions -->

```{r Missing Value Checker}
# Helper function to determine if any missing values
mv_checker <- function(df) {
  df %>% 
    map_df(~sum(is.na(.))) %>% 
    mutate(placeholder = 0) %>%
    pivot_longer(!placeholder, names_to="Column Names", values_to="Missing Values") %>% 
    select(-placeholder) %>% 
    arrange(desc(`Missing Values`)) %>% 
    mutate(`% of Total` = glue("{round(`Missing Values`/nrow(df) * 100, 2)}%"))
}
```

<!-- # Tibbles -->

<!-- ### Team Fixtures -->

```{r Download Team Fixtures and Combining}
# Downloading/Tidying the Fixtures data
afl_teams <- c("adelaide-crows", "brisbane-lions", "carlton", "collingwood", "essendon",
               "fremantle", "geelong-cats", "gold-coast-suns", "gws-giants", "hawthorn",
               "melbourne", "north-melbourne", "port-adelaide", "richmond", "st-kilda",
               "sydney-swans", "west-coast-eagles", "western-bulldogs")

# Downloads the 2021 afl fixtures for each team into the Fixtures_xlsx folder
download_fixtures <- function(home_path, fixtures_dir, afl_teams) {
  
  # Create directory to save files if not already exists
  if(!dir.exists(fixtures_dir)) {
    dir.create(fixtures_dir)
  }
  
  ### When downloading files there seems to be a glitch with the date/time column, so I did this manually.
  # Download files to directory: fixtures_dir 
  # if (length(list.files(fixtures_dir)) == 0) {
  #   setwd(paste0(home_path, "/", fixtures_dir))
  #   for (team in afl_teams) {
  #     download.file(url = paste0("https://fixturedownload.com/download/afl-2021-", team, "-UTC.xlsx"), 
  #                   destfile = paste0(team, "_fixtures.xlsx"),
  #                   method = "wget")
  #   }
  # }
}

download_fixtures(home_path, fixtures_dir, afl_teams)

# Appending each of the .xlsx files to form one .xlsx file with all team fixtures
merge_fixtures <- function(home_path, fixtures_name, fixtures_dir) {
  if (!file.exists(fixtures_name)) {
    setwd(paste0(home_path, "/", fixtures_dir))
    files <- list.files()
    fixtures <- list()
    for (i in 1:length(files)) {
      fixtures[[i]] <- read_excel(files[i], range=cell_cols("A:E")) %>% 
        mutate(across("Date", ~if_else(. < as.POSIXct("2021-04-04 03:00:00", tz="UTC"),
                                       . + hours(11),
                                       . + hours(10))),
               across("Date", ~format(., glue("%Y-%m-%d %I:%M")))) %>% 
        separate("Date", c("Date", "Kick-Off Time"), " ") %>% 
        mutate(across("Kick-Off Time", ~str_replace(., "^0", "")))
    }
    all_fixtures <<- reshape::merge_all(fixtures) %>% as_tibble()
    setwd(home_path)
    write_xlsx(all_fixtures, fixtures_name)
  } else {
    all_fixtures <<- read_excel(fixtures_name) # <<- means global variable
  }
}

merge_fixtures(home_path, fixtures_name, fixtures_dir)

# all_fixtures
```


<!-- ### AFL Ladder -->

```{r Ladder}
# Gathering and cleaning data on the AFL Ladder

# Using google chrome SelectorGadget to get the table ID
afl_ladder <- read_html("https://afltables.com/afl/seas/2021.html") %>% 
  html_element(".sortable") %>% 
  html_table()

# How to find the class name:
# afl_ladder %>% 
#   html_elements("table") %>% 
#   tail()
# Here it is the `2nd` last table

# Tidying the AFL Ladder
afl_ladder_tidy <- afl_ladder %>% 
  janitor::row_to_names(1) %>%
  slice(-19) %>% 
  rename(Position = `#`) %>% 
  mutate(across(c("W", "D", "L"), ~if_else(.=="", "0", .)),
         across(c("GF-BF", "GA-BA"), ~str_trunc(., 6, "right", "")),
         across(c("Position", "P", "W", "D", "L", "For", "Agn", "Pts"), ~as.integer(.)),
         across(c("GF-BF", "GA-BA", "%"), ~as.numeric(.)))

# afl_ladder_tidy
```

<!-- ### Historical Results -->

```{r Historical Results}
# Loading/cleaning Historical AFL results
hist_results <- read_excel("afl_historical_results.xlsx", 
                           sheet = "Data", 
                           range = "A2:L2338")

#mv_checker(hist_results)

hist_results_tidy <- hist_results %>% 
  rename("Kick-Off Time" = "Kick Off (local)") %>% 
  mutate(across("Date", ~as.Date(.)),
         across("Kick-Off Time", ~str_replace(format(., glue("%I:%M")), "^0", "")),
         across("Play Off Game?", ~replace_na(., "N")))

# hist_results_tidy
```

<!-- ### Player Statistics -->

<!-- What do the abbreviations mean? -->

<!-- -   TM - Team -->
<!-- -   GM - Games Played -->
<!-- -   KI - Kicks -->
<!-- -   MK - Marks -->
<!-- -   HB - Handballs -->
<!-- -   DI - Disposals -->
<!-- -   GL - Goals -->
<!-- -   BH - Behinds -->
<!-- -   HO - Hit Outs -->
<!-- -   TK - Tackles -->
<!-- -   RB - Rebound 50s -->
<!-- -   IF - Inside 50s -->
<!-- -   CL - Clearances -->
<!-- -   CG - Clangers -->
<!-- -   FF - Free Kicks For -->
<!-- -   FA - Free Kicks Against -->
<!-- -   BR - Brownlow Votes -->
<!-- -   CP - Contested Possessions -->
<!-- -   UP - Uncontested Possessions -->
<!-- -   CM - Contested Marks -->
<!-- -   MI - Marks Inside 50 -->
<!-- -   1% - One Percenters -->
<!-- -   BO - Bounces -->
<!-- -   GA - Goal Assists -->

```{r 2021 Player Statistics}
# AFL Player Statistics Data

# The link provided below is all the player statistics for rounds 1-5 2021.
player_stats <- read_html("https://afltables.com/afl/stats/2021a.html") %>% 
  html_element("table") %>% 
  html_table()

# All character cols have no missing values, great! We'll replace other cols with 0 since there all numeric
#mv_checker(player_stats)

# Fixing missing values
player_stats <- player_stats %>% 
  mutate(across(everything(), ~replace_na(., 0)))

# Creating a mapper from team abbreviation to full name
team_mapper <- list("WB" = "Western Bulldogs",
                    "HW" = "Hawthorn",
                    "ME" = "Melbourne",
                    "ES" = "Essendon",
                    "PA" = "Port Adelaide",
                    "CA" = "Carlton",
                    "GE" = "Geelong",
                    "AD" = "Adelaide",
                    "NM" = "North Melbourne",
                    "GW" = "Greater Western Sydney",
                    "CW" = "Collingwood",
                    "GC" = "Gold Coast",
                    "SK" = "St Kilda",
                    "RI" = "Richmond",
                    "BL" = "Brisbane",
                    "WC" = "West Coast",
                    "SY" = "Sydney",
                    "FR" = "Fremantle")

# Change team name from abbreviation to full name
change_team_name <- function(df, col, team_mapper) {
  for (i in seq_along(1:nrow(df))) {
    for (team in names(team_mapper)) {
      if (df[i, col] == team) {
        df[i, col] <- team_mapper[[team]]
      }
    }
  }
  return(df)
}

player_stats <- change_team_name(player_stats, "TM", team_mapper)

# Change players name from: "lname, fname" to "fname lname"
q <- player_stats
change_player_name <- function(df, col) {
  df[col] <- df %>% 
    pull({{ col }}) %>% 
    str_split(", ") %>% 
    map_chr(~paste(.[2], .[1]))
  return(df)
}

player_stats <- change_player_name(player_stats, "Player")

# player_stats
```

# Exploratory Data Analysis

### Leading Goal Scorers

```{r Top 10 leading scorers}
leading_goal_scorers <- player_stats %>% 
  select(Player, TM, GM, GL) %>% 
  arrange(desc(GL)) %>%
  slice(1:10) %>% 
  mutate(`GL / GM` = round(GL / GM, 2),
         across(Player, ~fct_reorder(., desc(GL))))

team_fill <- c("Western Bulldogs" = "#0373fc",
               "Hawthorn" = "#ffb303",
               "Melbourne" = "#10114a",
               "Essendon" = "#000000",
               "Port Adelaide" = "#03fcf4",
               "Carlton" = "#021716",
               "Geelong" = "#001645",
               "Adelaide" = "#ff2014",
               "North Melbourne" = "#2370cf",
               "Greater Western Sydney" = "#ff9900", 
               "Collingwood" = "#000000",
               "Gold Coast" = "#ff2600",
               "St Kilda" = "#000000",
               "Richmond" = "#fff200",
               "Brisbane" = "#701f03",
               "West Coast" = "#0c2ef0",
               "Sydney" = "#ff1900",
               "Fremantle" = "#6702ba")

top_10_scorers <- leading_goal_scorers %>% 
  ggplot() +
  geom_col(aes(Player, GL, fill=TM)) +
  labs(x = "Player Name",
       y = "Goals Scored",
       title = "Top 10 Leading Goal Scorers") +
  theme(axis.text.x = element_text(angle = 45, hjust=1),
        plot.title = element_text(hjust = 0.5)) +
  scale_fill_manual(values = team_fill, name = "Team Colors")

ggplotly(top_10_scorers)
```

### Leading Goal Assists

```{r Top 10 leaders most assists}
leading_most_goal_assists <- player_stats %>% 
  select(Player, TM, GM, GA) %>% 
  arrange(desc(GA)) %>%
  slice(1:10) %>% 
  mutate(`GA / GM` = round(GA / GM, 2),
         across(Player, ~fct_reorder(., desc(GA))))

top_10_goal_assists <- leading_most_goal_assists %>% 
  ggplot() +
  geom_col(aes(Player, GA, fill=TM)) +
  labs(x = "Player Name",
       y = "Goals Assists",
       title = "Top 10 Leaders With Most Assists") +
  theme(axis.text.x = element_text(angle = 45, hjust=1),
        plot.title = element_text(hjust = 0.5)) +
  scale_fill_manual(values = team_fill, name = "Team Colors")

ggplotly(top_10_goal_assists)
```


### Where are the Most Games Played at in 2021?

```{r}
most_popular_stadiums <- all_fixtures %>% 
  group_by(Location) %>% 
  count(sort=TRUE, name="Matches Played") %>% 
  ungroup() %>% 
  mutate(across(Location, ~fct_reorder(., desc(`Matches Played`))),
         State = c("VIC", "VIC", "SA", "WA", "QLD", "NSW", "QLD", "VIC",
                   "NSW", "TAS", "TAS", "ACT", "VIC", "QLD", "NT", "NT")) %>% 
  ggplot() +
  geom_col(aes(x=Location, y=`Matches Played`, fill=State)) +
  labs(title = "Most Popular Stadiums",
       x = "Stadium") +
  theme(axis.text.x = element_text(angle = 45, hjust=1),
        plot.title = element_text(hjust = 0.5))

ggplotly(most_popular_stadiums)
```

### Fixtures

Please note that from round 10 onwards the `Date` and `Kick-Off Time` are TBD as of writing. 

```{r}
shinyApp(

  ui = fluidPage(
    # Let step=0.5 to increment/decrement by 1, there seems to be some bug
    numericInput("roundNumber", "Round Number:", value=1, min=1, max=23, step=0.5),
    
    # List Fixtures
    tableOutput("2021FixtureInfo")
  ),

  server = function(input, output) {
    
    # Function allowing fixtures to be filtered by round
    all_fixtures_filtered <- function(df, round=NULL) {
      if (is.null(round)) {
          return(df)
      } else if (!is.null(round)) {
          df_filtered <- df %>% 
              filter(`Round Number` == round) %>% 
              select(-`Round Number`)
          return(df_filtered)
      }
    }
    
    # Update data base on changing round
    output$`2021FixtureInfo` <- renderTable({
        all_fixtures_filtered(all_fixtures, input$roundNumber)   
    })
  },

  options = list(height = 500)
)
```

### AFL Ladder

```{r}
afl_ladder_tidy %>% 
  datatable(rownames=FALSE)
```


